---
title: "Exercises: RMarkdown"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercises: RMarkdown

In this practical we will introduce the basics of using RMarkdown and some of its features.

## Introduction

R Markdown is a file format for making dynamic documents with R. An R Markdown document is written in markdown (an easy-to-write plain text format) and contains *chunks* of embedded R code.

An RMarkdown document is rendered by _knitting_ the file.

The `rmarkdown` package will call the `knitr` package. `knitr` will run each chunk of R code in the document and append the results of the code to the document next to the code chunk. This workflow saves time and facilitates reproducible reports.

Consider how authors typically include graphs (or tables, or numbers) in a report. The author makes the graph, saves it as a file, and then copy and pastes it into the final report. This process relies on manual labour. If the data changes, the author must repeat the entire process to update the graph.

In the R Markdown paradigm, each report contains the code it needs to make its own graphs, tables, numbers, etc. The author can automatically update the report by re-knitting.

You can manually render an R Markdown file with `rmarkdown::render()`.

## Creating a R Markdown document

To create an R Markdown report, open a plain text file and save it with the extension _.Rmd_. You can open a plain text file in your scripts editor by clicking _File > New File > Text File_ in the RStudio tool bar, or using the _Wizard_ by selecting _File > New File > R Markdown ..._

#### Rendering

To go from R Markdown to pdf or html you can click on the _Knit_ icon thats above the file in the scripts editor. A drop down menu will let you select the type of output that you want.
RStudio will show you a preview of the new output and save the output file in your working directory.


## Text formatting

Markdown is designed to be easy to read and easy to write. It is also very easy to learn. 

Go to the drop down menu and select _Help > Markdown Quck Reference_. This will open a cheat sheet in the Help pane.


## Code chunks

To run code inside an R Markdown document, you need to insert a chunk. There are three ways to do so:

* The keyboard shortcut Cmd/Ctrl + Alt + I

* The _Insert_ button icon in the editor tool bar

* By manually typing the chunk delimiters ```{r} and ```.

Think of a chunk like a function. A chunk should be relatively self-contained, and focussed around a single task.

### Chunk names
Chunks can be given an optional name: ```{r by-name}. This has three advantages:

* You can more easily navigate to specific chunks using the drop-down code navigator in the bottom-left of the script editor:

* Graphics produced by the chunks will have useful names that make them easier to use elsewhere. More on that in other important options.

* You can set up networks of cached chunks to avoid re-performing expensive computations on every run.

### Chunk options

Chunk output can be customised. A full list is here http://yihui.name/knitr/options/.

The most important set of options controls if your code block is executed and what results are inserted in the finished report:

* `eval = FALSE` prevents code from being evaluated. (And obviously if the code is not run, no results will be generated). This is useful for displaying example code, or for disabling a large block of code without commenting each line.

* `include = FALSE` runs the code, but doesn’t show the code or results in the final document. Use this for setup code that you don’t want cluttering your report.

* `echo = FALSE` prevents code, but not the results from appearing in the finished file. Use this when writing reports aimed at people who don’t want to see the underlying R code.

* `message = FALSE` or `warning = FALSE` prevents messages or warnings from appearing in the finished file.

* `results = 'hide'` hides printed output; `fig.show = 'hide'` hides plots.

* `error = TRUE` causes the render to continue even if code returns an error. This is rarely something you’ll want to include in the final version of your report, but can be very useful if you need to debug exactly what is going on inside your .Rmd. It’s also useful if you’re teaching R and want to deliberately include an error. The default, error = FALSE causes knitting to fail if there is a single error in the document.

## Tables in R Markdown

By default, R Markdown prints data frames and matrices as you’d see them in the console.
If you prefer that data be displayed with additional formatting you can use the `knitr::kable` function. For even deeper customisation, consider the `xtable`, `stargazer`, `pander`, `tables`, and `ascii` packages

## Inline code

To embed R code in a line of text, surround the code with a pair of back ticks and the letter r, like this.

```{r eval=F}
`r  ICER is 200/0.5`
```

knitr will replace the inline code with its result in your final document (inline code is always replaced by its result). The result will appear as if it were part of the original text

## YAML header

A YAML header to control how rmarkdown renders your _.Rmd_ file. A YAML header is a section of key: value pairs surrounded by --- marks.

The `output:` value determines what type of output to convert the file into when you call `rmarkdown::render()`.

`output:` recognizes the following values:

* `html_document`, which will create HTML output (default)
* `pdf_document`, which will create PDF output
* `word_document`, which will create Word output

If you use the RStudio IDE knit button to render your file, the selection you make in the gui will override the output: setting.

### Putting it all together in an example report

* Create a new R Markdown document with main sections with headings Introduction, Data, Methods, Results and Conclusions.

* Include some subheadings like Model Fitting and Recommendations, say.

* In the Conclusions make numbered list of the main findings.

* Include a link to the Briggs papers on which the original cost-effectiveness analysis is based i.e.
_An Introduction to Markov Modelling for Economic Evaluation, Andrew Briggs and Mark Sculpher_

* Change the date of the document so that it will always be today's date using the following

```{r eval=FALSE}
date: "`r format(Sys.time(), '%d %B, %Y')`"
```

Pick a date format that you prefer

* Include the plot created with the following code

```{r ceplane}
plot(10, 0.1,
     xlab = "Incremental QALYs",
     ylab = "Incremental costs",
     xlim = c(-1, 12),
     ylim = c(-0.1, 0.2),
     pch = 20)
abline(a = 0, b = 0.02)
```

* Resize this plot using the chunk options

* Create a table that shows the cost and QALYs for two interventions and their incremental values and ICER. Here something to start with.

intervention | cost | QALY | delta c | delta QALY | ICER
-------------|------|------|---------|------------|-----

* Now create the same table but do this programmatically. That is create a data frame with the entries and then use `knitr::kable` to convert to markdown.

Here's an example

```{r eval=FALSE}
data.frame(intervention = c("drug a", "drug b"),
           cost = c(100, 50),
           QALY = c(20, 10)) |> 
knitr::kable()
```

