---
title: "Debugging in R"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Debugging in R

## Introduction

In this practical we will introduce and practise basic debugging in R and RStudio.
We will use the cost-effectiveness Markov model code to demonstrate this.


Open up the scripts `debugging_script.R` run it.

This will first source the files `ce_markov_debug1.R` and `p_matrix_cycle.R` by running

```{r}
source("ce_markov_debug1.R")
```

Make sure this is in the correct folder for you.

This is the main script from which to run the analysis which calls our functions.
You should get the following error

```
Error in cycle_state_costs[i, ] <- (state_c_matrix[treatment = i, ] %*%  : 
  number of items to replace is not a multiple of replacement length
```

Lets investigate...

## Using `debug()`

The functions `browser()` and `debug()` are simple to use and very powerful at identifying the source of an error.

`debug()` works by telling R to enter the debugger at the first line whenever a given function is called.

```{r}
debug(ce_markov)
```

This breakpoint on the first line will remain until you remove it with
 
```{r}
undebug(ce_markov)
```

Sometimes you just want to run the debugger a single time and so instead use

```{r}
debugonce(ce_markov)
```

Now run

```{r eval=FALSE}
ce_res <- ce_markov(start_pop = c(n_pop, 0, 0),
                    p_matrix,
                    state_c_matrix,
                    trans_c_matrix,
                    state_q_matrix)
```

The GUI will show you the `ce_markov()` script with the body of the function highlighted.
A green arrow on the left hand side will point at the first line of the code. This indicated where the debugger has stopped at and hasn't executed any of the lines below it.

Execute the next line of code by either pressing `n` or `Enter` or _F10_ or clicking on the _Next_ button.

Notice in the Environment pane that the values are for those passed to the function.

If you 'step through' the code by pressing _Next_ a few more times you will see the Environment variable being up dated as the highlighted code moved down line by line.

When you get to the `for (j in 2:n_cycles)` loop you'll notice that pressing _Next_ takes you around every iteration. If you want to fast forward and leave the loop having run all iterations in one step then you can "step out" of the loop with _Shift+F7_.

__Which line does the error occur at?__

The `debug()` function is pretty cool because it can also show you the insides of other people's functions.

For example, try

```{r eval=FALSE}
debugonce(mean)
mean(c(1,2,"three"))
```

This will send you to the `mean.default` method

```{r eval=FALSE}
function (x, trim = 0, na.rm = FALSE, ...) 
{
  if (!is.numeric(x) && !is.complex(x) && !is.logical(x)) {
    warning("argument is not numeric or logical: returning NA")
    return(NA_real_)
  }
  if (isTRUE(na.rm)) 
    x <- x[!is.na(x)]
  if (!is.numeric(trim) || length(trim) != 1L) 
    stop("'trim' must be numeric of length one")
  n <- length(x)
  if (trim > 0 && n) {
    if (is.complex(x)) 
      stop("trimmed means are not defined for complex data")
    if (anyNA(x)) 
      return(NA_real_)
    if (trim >= 0.5) 
      return(stats::median(x, na.rm = FALSE))
    lo <- floor(n * trim) + 1
    hi <- n + 1 - lo
    x <- sort.int(x, partial = unique(c(lo, hi)))[lo:hi]
  }
  .Internal(mean(x))
}
```

__Have a look inside some other functions in the same way.__


## Using `browser()`

Every time you use `debug()` it starts at the first line of the function.
Now that we have identified where the error is thrown, or if you had an idea beforehand anyway, you don't want to step through all the code to get to the interesting bit.
In this case you can use the browser()` function.

Simply write `browser()` on the line where you want the break point to be.
Include `browser()` in a sensible place in `ce_markov()` and rerun the function.
Once the break point has been activated you can step through the code in the same way as before.
Don't forget to re-source the function to tell R that you have included new code.

#### Editor break points

A further way that you can include break points is interactively by left clicking the margin of the Editor pane to the left of the row numbers. You'll see that a red dot appears and this is equivalent to the `browser()` function at this line.

You can have multiple break points through out the code and when you continue running the code after one break point (pressing `c` or _Shift+F5_) then the code execution stops at the next one.

__Try using multiple break points, setting them via the Editor__


#### Stepping _in to_ functions

Run `ce_markov()` again with `debugonce()`. We may have a suspicion that `p_matrix_cycle` is responsible for something going wrong. We could add multiple break point as above with one inside of `p_matrix_cycle`. Alternatively, when we get to this line of code in `ce_markov()` we can "step in to" the function pressing `s` or _Shift+F4_.
This is like taking us one layer down into the code.

__Try this stepping in to `p_matrix_cycle` from `ce_markov`. You can also step in to base R functions.__

## Conditional break points

As was the case against using `debug()` we don't want to be stepping through loads of code to get to the point where the error occurs or where we think that the problem originated in the code.

We can further refine our use of break points by only entering the debugger when a certain condition is met. This is commonly when a loop is at a certain (large number) of iterations. 

It will look something like this

```{r eval=FALSE}
if <CONDITIONAL STATEMENT> browser()
```

__Include a conditional breakpoint in the `ce_markov()` function for when the number of cycles is over 20__



## But where is the bug?

If we go back to the original error message

```
Error in cycle_state_costs[i, ] <- (state_c_matrix[treatment = i, ] %*%  : 
  number of items to replace is not a multiple of replacement length
```

This is telling us that the dimensions of `cycle_state_costs` and the result of the right hand side don't match.

Let us first see what these dimensions are. Add a break point just before this line and then type

```{r eval=FALSE}
dim(state_c_matrix[treatment = i, ] %*% pop[, , treatment = i])
dim(cycle_state_costs[i, ])
cycle_state_costs[i, ]
```

It appears that the object of length 2 is being assigned 46 entries.
If we look at the full matrix

```{r eval=FALSE}
cycle_state_costs
```

we see that this has 46 rows which is a remarkable coincidence.
We can hypothesise that the rows and columns have gotten mixed up somewhere along the way earlier in the code and the error is thrown later on when we try to use these objects.

If we search for `cycle_state_costs` we see that it is initiated with `cycle_empty_array`.
This is the culprit. We can see that the `dim` argument has the order cycle, treatment but the `dimnames` argument has the order treatment, cycle.
Swapping `dim = c(n_treat, n_cycles)` should fix the error.
This is a relatively easy bug to fix because we can spot the inconsistency in the code. Some bugs can be particularly difficult to catch but using the debugger to hone in and probe inside of the code can be invaluable.



